A Way Forward in Parallelising Dynamic Languages

### 2014-06-17 talk/icooolps2014/position-paper.tex
some adjustments

\begin{abstract}
  Dynamic languages became very popular in recent years. At some
  point, the need for concurrency arose, and many of them made the
  choice to use a single global interpreter lock (GIL) to synchronise
  the interpreter in a multithreading scenario. This choice however
  makes it impossible to actually run code in parallel.

@@

\section{Introduction}
In a world where computers get more and more cores and single-thread
performance increases less and less every year, many dynamic languages
have a problem. While there is certainly a lot of popularity around
languages like Python and Ruby, their ability to make use of multiple
cores is somewhat limited. For ease of implementation they chose to
use a single, global interpreter lock (GIL) to synchronise the
execution of code in multiple threads. While this is a
straight-forward way to eliminate synchronisation issues in the
interpreter, it prevents parallel execution. Code executed in multiple
threads will be serialised over this GIL so that only one thread can
execute at a time.

@@

These requirements are not easy to meet. We argue that STM is the
overall winner. While it currently has a big performance problem, it
gets more points in all the other categories. We think that it is the
only solution that also provides a better synchronisation mechanism to
the application in the form of parallelisable atomic blocks.

%% \subsection{Issue}
%% The issue that we want to discuss is how to efficiently support
%% multi-core parallel execution of code in dynamic languages that were
%% designed with GIL semantics in mind.

%% Furthermore, a solution to this problem should also bring better
%% synchronisation mechanism with it...

%% (supporting (large) atomic blocks for synchronisation)

%% \subsection{Our Position}
%% Current solutions for replacing the GIL include STM, HTM, and
%% fine-grained locking. STM is usually too slow, HTM very limited, and
%% locking suffers from complexity that makes it hard to implement
%% correctly. We argue that the best way forward is still STM and that
%% its performance problem can be solved.

%% Current solutions like STM, HTM, and fine-grained locking are slow, hard
%% to implement correctly, and don't fit the specific problems of dynamic
%% language.  STM is the best way forward but has bad performance, so we
%% fix that.

@@

In this section we examine the approaches and highlight their
advantages and disadvantages.
%% \paragraph{dynamic language VM problems}
%% XXX:
%% - high allocation rate (short lived objects)\\
%% - (don't know anything about the program that runs until it actually runs: arbitrary atomic block size)

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions\footnote{This
also applies to Abstract Syntax Tree (AST) interpreters, where the GIL
may only be released between interpreting two AST nodes.}. The interpreter
can thus rely on complete isolation and atomicity for the
instructions' execution. Thus, accesses to data structures like
dictionaries and lists happen atomically and do not need additional
protection from data races when shared between threads.

The GIL also provides the application with a sequential consistency
model~\cite{lamport79}. This can be very valuable as it means less
surprises for the programmer. For example in Table~\ref{tab:seq_cons},
the programmer can expect the critical section to only be entered by
one thread.  If the model allowed to buffer the writes, both threads
may enter the critical section at the same time.

\begin{table}[!ht]
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      Thread 1                      & Thread 2                      \\
      \hline
      \multicolumn{2}{|l|}{\texttt{A = B = 0}}                      \\
      \hline
      \texttt{A = 1}                & \texttt{B = 1}                \\
      \texttt{if B == 0:}           & \texttt{if A == 0:}           \\
      \texttt{    critical section} & \texttt{    critical section} \\
      \hline
    \end{tabular}
    \caption{Critical section with a sequential consistency model.}
    \label{tab:seq_cons}
  \end{center}
\end{table}

@@

The GIL also allows for easy integration with external C libraries that
may not be thread-safe. For the duration of the calls, we
simply do not release the GIL. External libraries that are explicitly
thread-safe can voluntarily release the GIL themselves in order to
still provide some parallelism. This is done for example for
potentially long I/O operations. Consequently, I/O-bound,
multithreaded applications can actually parallelise to some
degree. Again, a potential solution should be able to integrate with
external libraries with similar ease. We will however focus our
argumentation more on running code in the interpreted language in
parallel, not the external C calls.

@@

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. Interpreters using a GIL can simply guarantee
that the GIL is not released during the execution of the atomic
block. Of course, this still means that no two atomic blocks can
execute in parallel or even concurrently.  Potential solutions are
preferable if they provide a good way to implement atomic blocks that
are also able to be executed in parallel.

@@

For the discussion we define a set of criteria to evaluate the
multiple potential solutions for removing or avoiding the GIL and its
limitations:

@@

The first obvious candidate to replace the GIL is to use multiple
locks instead of a single global lock. By refining the granularity of
the locking approach, we gain the ability to run code that does not
access the same objects in parallel. What we lose instead is the
simplicity of the GIL approach. With every additional lock, the
likeliness of deadlocks grows, as well as the overhead that acquiring
and releasing locks produces. The former means that sometimes it is
necessary to fall back to less fine-grained locking, preventing some
potential parallelism in order to keep the complexity manageable.
The latter means that we lose a bit of performance in the
single-threaded case compared to the GIL, which requires much less
acquire-release operations.

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the Java Virtual Machine (JVM) and
that uses fine-grained locking\footnote{The performance impact of
fine-grained locking is milder in Java than it would be in a typical piece
of C code; see e.g.~\cite{biased}.} to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks -- every dictionary, list, instance, or mutable
object in general needs a lock. Compared to e.g.\ Java, object
attributes are backed by a dictionary. Accesses to it must be
synchronised because the interpreter could crash otherwise.  Since
there is no central location for all these locks, the
complexity of the implementation is quite a bit larger compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One could simply use one lock per library
to avoid this issue.

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes\footnote{There are rare cases where not having atomic
  bytecodes actually changes the semantics. E.g.\ in Jython,
  \texttt{dict1.update(dict2)} is not atomic: it first reads data from
  \texttt{dict2} with \texttt{dict2}'s lock, and then puts it into
  \texttt{dict1} with \texttt{dict1}'s lock. A lot can happen
  in-between.}. An implementation has to follow the GIL semantics very
closely, otherwise it may expose some latent data races in existing
applications which are just not exposed with a GIL. This approach does
however not provide a better parallelising synchronisation mechanism
to the application like e.g.\ atomic blocks.

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share anything, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. At the extreme, there
are applications that parallelise perfectly simply by running
independent processes; some web servers and some numeric computations
do.

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$ overhead~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
reasonable in general to expose the hardware-transactions to the
application in the form of atomic blocks, because doing so would
require the system to support much longer transactions.

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The points were assigned
according to the criteria described in \ref{sec:pot_solutions}. Since
the criteria are defined intuitively, there are no formal justifications
for the assigned points. The reader is thus advised to take them with a
grain of salt.

@@

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on a STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case which gives us an
advantage over other STM systems that are more general. With this
approach, initial results suggest that we can keep the overhead of STM
well below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too.

>>>>

\begin{abstract}
  Dynamic languages became very popular in recent years. At some
  point, the need for concurrency arose, and many of them made the
  choice to use a single global interpreter lock (GIL) to synchronise
  the interpreter in a multithreading scenario. This choice, however,
  makes it impossible to actually run code in parallel.

@@

\section{Introduction}
In a world where computers get more and more cores and single-thread
performance increases less and less every year, many dynamic languages
have a problem. While there is certainly a lot of popularity around
languages like Python and Ruby, their ability to make use of multiple
cores is somewhat limited. For ease of implementation, they chose to
use a single, global interpreter lock (GIL) to synchronise the
execution of code in multiple threads. While this is a
straight-forward way to eliminate synchronisation issues in the
interpreter, it prevents parallel execution. Code executed in multiple
threads will be serialised over this GIL so that only one thread can
execute at a time.

@@

These requirements are not easy to meet. The author's position is that
STM provides the best way forward. While STM currently has a big
performance problem, it gets more points in the other categories. We
think that it is the only solution that also provides a better
synchronisation mechanism to the application in the form of
parallelisable atomic blocks.  In the following section, we try to
present a balanced view of the compared approaches.

@@

In this section we first explain the motivation for using a GIL and
then examine different approaches to remove or avoid it -- highlighting
their advantages and disadvantages.

\subsection{Why is there a GIL?}
The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions\footnote{This
also applies to Abstract Syntax Tree (AST) interpreters, where the GIL
may only be released between interpreting two AST nodes.}. The interpreter
can thus rely on complete isolation and atomicity for the
instructions' execution. Also, accesses to data structures like
dictionaries and lists happen atomically and do not need additional
protection from data races when shared between threads.

The GIL also provides the application with a sequential consistency
model~\cite{lamport79}. This can be very valuable as it means less
surprises for the programmer. For example in Table~\ref{tab:seq_cons},
the programmer can expect the critical section to only be entered by
one thread.  On the other hand, if the model allowed to buffer the
writes, both threads may enter the critical section at the same time.

\begin{table}[!ht]
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      Thread 1                      & Thread 2                      \\
      \hline
      \multicolumn{2}{|c|}{\texttt{A = B = 0}}                      \\
      \hline
      \texttt{A = 1}                & \texttt{B = 1}                \\
      \texttt{if B == 0:}           & \texttt{if A == 0:}           \\
      \multicolumn{2}{|c|}{only one thread enters here}             \\
      \multicolumn{2}{|c|}{(e.g.\ critical section)}             \\
      \hline
    \end{tabular}
    \caption{Critical section with a sequential consistency model.}
    \label{tab:seq_cons}
  \end{center}
\end{table}

@@

The GIL also allows for easy integration with external C libraries that
may not be thread-safe. For the duration of the calls, we
simply do not release the GIL. External libraries that are explicitly
thread-safe can voluntarily release the GIL themselves in order to
still provide some parallelism. This is done for example for
potentially long I/O operations. Consequently, I/O-bound,
multithreaded applications can actually parallelise to some
degree. Again, a potential solution should be able to integrate with
external libraries with similar ease. We will however focus our
argumentation more on running code in the interpreted language in
parallel, not the external C code.

@@

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. Interpreters using a GIL can simply guarantee
that the GIL is not released during the execution of the atomic
block. Of course, this still means that no two atomic blocks can
execute in parallel or even concurrently.  Potential solutions are
preferable if they provide a good way to implement atomic blocks
(or another, comparable synchronisation mechanism) that
are also able to be executed in parallel.

@@

For the discussion, we define a set of criteria to evaluate the
potential solutions for removing or avoiding the GIL and its
limitations:

@@

The first obvious candidate to replace the GIL is to use multiple
locks instead of a single global lock. By refining the granularity of
the locking approach, we gain the ability to run code that does not
access the same objects in parallel. What we lose instead is the
simplicity of the GIL approach. With every additional lock, the
likeliness of deadlocks grows, as well as the overhead that acquiring
and releasing locks produces. The former means that sometimes it is
necessary to fall back to less fine-grained locking, preventing some
potential parallelism in order to keep the complexity manageable.
The latter means that we lose a bit of performance compared to the
GIL, which requires much less acquire-release operations.

Jython~\cite{webjython} is one project that implements an
interpreter for Python on the Java Virtual Machine (JVM) and
that uses fine-grained locking\footnote{The performance impact of
fine-grained locking is milder on the JVM than it would be in a typical piece
of C code; see e.g.~\cite{biased}.} to correctly synchronise the
interpreter. For a language like Python, one needs quite a few,
carefully placed locks -- every dictionary, list, instance, or mutable
object in general needs a lock. Compared to e.g.\ Java, object
attributes are backed by a dictionary. Accesses to it must be
synchronised because the interpreter could crash otherwise.  Since
there is no central location for all these locks, the
complexity of the implementation is quite a bit larger compared to
using a GIL. Integrating external, non-thread-safe libraries should
however be very simple too. One can simply use one lock per library
to avoid this issue.

In the end, fine-grained locking can transparently replace the GIL and
therefore parallelise existing applications, generally without any
changes. An implementation has to follow the GIL semantics very
closely, otherwise it may expose some latent data races in existing
applications which are just not exposed with a GIL\footnote{There are
  rare cases where not having atomic bytecodes actually changes the
  semantics. E.g.\ in Jython, \texttt{dict1.update(dict2)} is not
  atomic: it first reads data from \texttt{dict2} with \texttt{dict2}'s
  lock, and then puts it into \texttt{dict1} with \texttt{dict1}'s
  lock. A lot can happen in-between.}. This approach does however not
provide a better parallelising synchronisation mechanism to the
application and still requires explicit locking in the application.

@@

There are also approaches that work around the GIL instead of trying
to replace it. If an application can be split into completely
independent parts that only very rarely need to share something, or
only do so via an external program like a database, then it is
sensible to have one GIL per independent part. At the extreme, there
are applications that parallelise perfectly simply by running
independent processes; some web servers and some numeric computations
do.

@@

The performance of HTM is pretty good as it does not introduce much
overhead ($<40\%$~\cite{odaira14}). And it can transparently
parallelise existing applications to some degree. The implementation
is very straight-forward because it directly replaces the GIL in a
central place. HTM is also directly compatible with any external
library that needs to be integrated and synchronised for use in
multiple threads. The one thing that is missing is support for a
better synchronisation mechanism for the application. It is not
reasonable in general to expose the hardware-transactions to the
application in the form of atomic blocks, because doing so would
require the system to support much longer transactions.

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The points were assigned
according to the criteria described in \ref{sec:pot_solutions}. Since
the criteria are defined intuitively, there are no formal justifications
for the number of points. The reader is thus advised to take the result
with a grain of salt.

@@

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on an STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case, which gives us an
advantage over other STM systems that are more general. With this
approach, initial results suggest that we can keep the overhead of STM
well below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too.
